Interim report 2\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}

### 2014-09-01 interimreport2.tex
Draft final revised

This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes. Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two: NZDep 1--5 and NZDep 6--10. We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category. By doing so, we reduce the zeroinflation present in the data.

The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see "None", Table~\ref{table:case}).

@@

The numbers of cases per category for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}. Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses.

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-010}
\end{center}
\caption{Per capita rates of measles infections broken down by ethicity, see Table~\ref{table:percap} for details}
\label{fig:percap}
\end{figure}

\begin{table}\tiny
\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
%latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllrrr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Population}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per Capita}\tabularnewline
\hline
1-5&0-2&Asian&$   6094$&$ 11$&$0.0018$\tabularnewline
6-10&0-2&Asian&$   6806$&$  8$&$0.0012$\tabularnewline
1-5&3-5&Asian&$   6094$&$  1$&$0.0002$\tabularnewline
6-10&3-5&Asian&$   6806$&$  0$&$0.0000$\tabularnewline
1-5&6-17&Asian&$  33918$&$ 11$&$0.0003$\tabularnewline
6-10&6-17&Asian&$  28905$&$  5$&$0.0002$\tabularnewline
1-5&18-24&Asian&$  22917$&$  3$&$0.0001$\tabularnewline
6-10&18-24&Asian&$  34107$&$  5$&$0.0001$\tabularnewline
1-5&25+&Asian&$  96357$&$ 10$&$0.0001$\tabularnewline
6-10&25+&Asian&$  98715$&$ 13$&$0.0001$\tabularnewline
1-5&0-2&European&$  57872$&$ 83$&$0.0014$\tabularnewline
6-10&0-2&European&$  45445$&$ 64$&$0.0014$\tabularnewline
1-5&3-5&European&$  57872$&$ 42$&$0.0007$\tabularnewline
6-10&3-5&European&$  45445$&$ 17$&$0.0004$\tabularnewline
1-5&6-17&European&$ 264330$&$219$&$0.0008$\tabularnewline
6-10&6-17&European&$ 182937$&$ 80$&$0.0004$\tabularnewline
1-5&18-24&European&$ 107649$&$ 34$&$0.0003$\tabularnewline
6-10&18-24&European&$ 117840$&$ 36$&$0.0003$\tabularnewline
1-5&25+&European&$1001916$&$ 78$&$0.0001$\tabularnewline
6-10&25+&European&$ 724317$&$ 51$&$0.0001$\tabularnewline
1-5&0-2&Maori&$  10003$&$ 18$&$0.0018$\tabularnewline
6-10&0-2&Maori&$  30104$&$ 48$&$0.0016$\tabularnewline
1-5&3-5&Maori&$  10003$&$  7$&$0.0007$\tabularnewline
6-10&3-5&Maori&$  30104$&$ 11$&$0.0004$\tabularnewline
1-5&6-17&Maori&$  40461$&$ 19$&$0.0005$\tabularnewline
6-10&6-17&Maori&$ 116640$&$ 92$&$0.0008$\tabularnewline
1-5&18-24&Maori&$  15360$&$  5$&$0.0003$\tabularnewline
6-10&18-24&Maori&$  48495$&$  8$&$0.0002$\tabularnewline
1-5&25+&Maori&$  71217$&$  2$&$0.0000$\tabularnewline
6-10&25+&Maori&$ 192729$&$  6$&$0.0000$\tabularnewline
1-5&0-2&MLA&$    728$&$  0$&$0.0000$\tabularnewline
6-10&0-2&MLA&$   1290$&$  3$&$0.0023$\tabularnewline
1-5&3-5&MLA&$    728$&$  0$&$0.0000$\tabularnewline
6-10&3-5&MLA&$   1290$&$  1$&$0.0008$\tabularnewline
1-5&6-17&MLA&$   2991$&$  0$&$0.0000$\tabularnewline
6-10&6-17&MLA&$   4539$&$  1$&$0.0002$\tabularnewline
1-5&18-24&MLA&$   1710$&$  0$&$0.0000$\tabularnewline
6-10&18-24&MLA&$   3078$&$  2$&$0.0006$\tabularnewline
1-5&25+&MLA&$   8028$&$  0$&$0.0000$\tabularnewline
6-10&25+&MLA&$  10335$&$  6$&$0.0006$\tabularnewline
1-5&0-2&Pacific&$   2093$&$  5$&$0.0024$\tabularnewline
6-10&0-2&Pacific&$  13124$&$ 58$&$0.0044$\tabularnewline
1-5&3-5&Pacific&$   2093$&$  0$&$0.0000$\tabularnewline
6-10&3-5&Pacific&$  13124$&$  3$&$0.0002$\tabularnewline
1-5&6-17&Pacific&$   8541$&$  5$&$0.0006$\tabularnewline
6-10&6-17&Pacific&$  51183$&$ 22$&$0.0004$\tabularnewline
1-5&18-24&Pacific&$   3972$&$  1$&$0.0003$\tabularnewline
6-10&18-24&Pacific&$  22098$&$  8$&$0.0004$\tabularnewline
1-5&25+&Pacific&$  18492$&$  2$&$0.0001$\tabularnewline
6-10&25+&Pacific&$  91533$&$ 11$&$0.0001$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:percap}
\end{table}

However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category, and along with small population sizes (Table\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table\ref{table:percap}).

For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used and present the results of a quasipoission regression model. We also account for differences in population sizes by using an offset term, the log(population size). We used a model simplification approach, by beginning our analyses with all terms and all interactions, and simplifying the models through removal of non-significant higher order interaction terms. Thus, the final model that remained with all significant interaction terms had the following linear predictor:

\begin{equation} \label{eq:reg}
 log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a_e(x_a * x_e)+ log(population)  + \epsilon
  \end{equation}
Where $\alpha$ is the intercept, $y$ cases, $_a$ age, $_e$ Prioritized Ethnicity, $_n$ NZDep, and $\epsilon$ the error term.

@@

The distribution of the cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-014}
\end{center}
\caption{Histogram of cases per category used in the final regression model}
\label{fig:histcasecat}
\end{figure}

The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-015}
\end{center}
\caption{Regression model predictions plotted against the cases (Equation~\ref{eq:reg})}
\label{fig:predict}
\end{figure}

@@

A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}. The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented, though herd immunity of approximately 95 percent a homogeneously mixed immune population is recommended to interupt measles \citep{roberts0,roberts4}. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles. Thus, the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, as etimated using the NZ census data \citep{stats14} and the proportion of each age class immune, based on data provided to us by the Ministry of Health on commencing this project. The figure shown (Figure~\ref{fig:naive}) and for all subsequent analyses assume a currently susceptible population in New Zealand of approximately 11\% of the total population.

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-020}
\end{center}
\caption{New Zealand population by age and estimates numbers of naive people in each age class}
\label{fig:naive}
\end{figure}

@@

\begin{equation} \label{eq:pc}
p_c = 1-(1/R_0)
  \end{equation}

We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$. We estimated the susceptible population in NewZealand from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}). Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.

We then used the proportions of the population that were estimated to require vaccination (Equation~\ref{eq:pc}), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals. These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the benefit--cost analyses section below (Section~\ref{sub:cost_benefit}).

finally, we used the same model structure and generation times to estimate the epidemic sizes once vaccination catch up protocols had been implemented assuming $R_v$ < 1 and the population of suscpetible people in New Zealand was now reduced by the numbers vaccinated. Because the model is stochastic, we ran 1000 simulations and report the mean values, though the distribution of these is shown in the benefit--cost section (Section~\ref{sub:cost_benefit}).

@@

\begin{table}
\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
%latex.default(costtable3, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lrrr}
\hline\hline
\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Admitted}&\multicolumn{1}{c}{Proportion}\tabularnewline
\hline
\textless  15 months&$ 21$&$10$&$0.47$\tabularnewline
15 months – 3 years&$  7$&$ 1$&$0.14$\tabularnewline
4 – 9 years&$  8$&$ 0$&$0.00$\tabularnewline
10 -1 19 years&$132$&$12$&$0.09$\tabularnewline
\textgreater  19 years&$ 33$&$11$&$0.33$\tabularnewline
Total&$201$&$34$&$0.17$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:freq}
\end{table}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-028}
\end{center}
\caption{Number of cases attending hospital and stay duration}
\label{fig:hosp}
\end{figure}

@@

\begin{table}
\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2000--2014}
%latex.default(costtable4, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllrrl}
\hline\hline
\multicolumn{1}{c}{Year}&\multicolumn{1}{c}{Gender}&\multicolumn{1}{c}{Cost}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Length.of.stay}&\multicolumn{1}{c}{Cost.per.case}\tabularnewline
\hline
2000&F&4,296&$  2$&$  4$&2,148\tabularnewline
&M&4,554&$  4$&$  9$&1,139\tabularnewline
&Total&8,850&$  6$&$ 13$&1,475\tabularnewline
2001&F&3,740&$  5$&$  5$&748\tabularnewline
&M&7,527&$  8$&$ 13$&941\tabularnewline
&Total&11,267&$ 13$&$ 18$&867\tabularnewline
2002&F&924&$  2$&$  0$&462\tabularnewline
&M&2,945&$  3$&$  2$&982\tabularnewline
&Total&3,869&$  5$&$  2$&774\tabularnewline
2003&F&9,766&$  8$&$ 12$&1,221\tabularnewline
&M&475&$  1$&$  0$&475\tabularnewline
&Total&10,241&$  9$&$ 12$&1,138\tabularnewline
2004&F&1,437&$  1$&$  2$&1,437\tabularnewline
&M&3,328&$  3$&$  3$&1,109\tabularnewline
&Total&4,765&$  4$&$  5$&1,191\tabularnewline
2005&F&0&$  0$&$  0$&0\tabularnewline
&M&5,111&$  3$&$ 11$&1,704\tabularnewline
&Total&5,111&$  3$&$ 11$&1,704\tabularnewline
2006&F&0&$  0$&$  0$&0\tabularnewline
&M&602&$  1$&$  0$&602\tabularnewline
&Total&602&$  1$&$  0$&602\tabularnewline
2007&F&1,930&$  1$&$  3$&1,930\tabularnewline
&M&81,046&$  4$&$ 22$&20,262\tabularnewline
&Total&82,977&$  5$&$ 25$&16,595\tabularnewline
2008&F&714&$  1$&$  0$&714\tabularnewline
&M&2,324&$  2$&$  1$&1,162\tabularnewline
                    &Total&3,038&$  3$&$  1$&1,013\tabularnewline
2009&F&11,953&$  7$&$ 15$&1,708\tabularnewline
&M&28,830&$ 22$&$ 23$&1,310\tabularnewline
                    &Total&40,782&$ 29$&$ 38$&1,406\tabularnewline
2010&F&5,884&$  4$&$  5$&1,471\tabularnewline
&M&817&$  1$&$  0$&817\tabularnewline
                    &Total&6,701&$  5$&$  5$&1,340\tabularnewline
2011&F&103,460&$ 66$&$ 86$&1,568\tabularnewline
&M&101,842&$ 66$&$103$&1,543\tabularnewline
                    &Total&205,303&$132$&$189$&1,555\tabularnewline
2012&F&13,054&$  8$&$  6$&1,632\tabularnewline
&M&15,486&$ 11$&$  6$&1,408\tabularnewline
                    &Total&28,540&$ 19$&$ 12$&1,502\tabularnewline
2013&F&1,800&$  1$&$  2$&1,800\tabularnewline
&M&3,530&$  3$&$  4$&1,177\tabularnewline
                    &Total&5,330&$  4$&$  6$&1,333\tabularnewline
2014&F&55,633&$ 21$&$ 46$&2,649\tabularnewline
&M&77,014&$ 34$&$ 87$&2,265\tabularnewline
&Total&132,647&$ 55$&$133$&2,412\tabularnewline
2000-2014&      F&335,431&$166$&$284$&2,021\tabularnewline
&     M&214,591&$127$&$186$&1,690\tabularnewline
&   TOTAL&550,022&$293$&$470$&1,877\tabularnewline
\hline
\end{tabular}\end{center}\label{table:cases}
 \centering
 \begin{tablenotes}
      \small
      \item As of 7 August, 2014.
    \end{tablenotes}
\end{table}

@@

\subsection{Benefit--cost analyses methods}
To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{sec:epidemic_modelling}), we did several things. Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ from (Section~\ref{sec:epidemic_modelling}) in the estimated susceptible population of naive New Zealanders (Figure~\ref{fig:naive}), assuming recovery from infection led to immunity and thus constantly reducing the population size by the number recovered. We used these values of numbers of predicted cases and the cost figures above to estimate what the cost of not vaccinating additional populations would be. We also simulated what we might expect measles outbreaks to look like following introductions in the population, given that $R_v$ would be one, though where our estimates for $R_v$ are already < 1, we use this. We can therefore estimate the number of cases prevented, and the savings made from the additional vaccinations. We then use the costs of the catch up vaccination schemes, the costs of the expected measles-related costs due to constant introduction of measles despite increased population immunity, and the savings from reduced measles cases to work out the benefit--cost ratio ((B/C), where a B/C ratio > 1 means that the program benefits exceed their costs. A B/C value less than one suggests the costs are higher than the economic benefits. Lastly, benefits were assessed over a 10--year time period, using a discounting rate of 3\% discount per year for the costs, as is common for healthcare discounting \citep{honeycutt6}:

>>>>

This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes. Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to two: NZDep 1--5 and NZDep 6--10. We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category. By doing so, we reduce the zeroinflation present in the data.

The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories. For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:case}).

@@

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}. Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses. However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category, and along with small population sizes (Table~\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\ref{table:percap}).

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-010}
\end{center}
\caption{Per capita rates of measles infections broken down by ethnicity, see Table~\ref{table:percap} for details}
\label{fig:percap}
\end{figure}

\begin{table}\tiny
\caption{Numbers of measles cases, population sizes and per capita rates of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
%latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllrrr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Population}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per capita}\tabularnewline
\hline
1-5&0-2&Asian&$   6094$&$ 11$&$0.0018$\tabularnewline
6-10&0-2&Asian&$   6806$&$  8$&$0.0012$\tabularnewline
1-5&3-5&Asian&$   6094$&$  1$&$0.0002$\tabularnewline
6-10&3-5&Asian&$   6806$&$  0$&$0.0000$\tabularnewline
1-5&6-17&Asian&$  33918$&$ 11$&$0.0003$\tabularnewline
6-10&6-17&Asian&$  28905$&$  5$&$0.0002$\tabularnewline
1-5&18-24&Asian&$  22917$&$  3$&$0.0001$\tabularnewline
6-10&18-24&Asian&$  34107$&$  5$&$0.0001$\tabularnewline
1-5&25+&Asian&$  96357$&$ 10$&$0.0001$\tabularnewline
6-10&25+&Asian&$  98715$&$ 13$&$0.0001$\tabularnewline
1-5&0-2&European&$  57872$&$ 83$&$0.0014$\tabularnewline
6-10&0-2&European&$  45445$&$ 64$&$0.0014$\tabularnewline
1-5&3-5&European&$  57872$&$ 42$&$0.0007$\tabularnewline
6-10&3-5&European&$  45445$&$ 17$&$0.0004$\tabularnewline
1-5&6-17&European&$ 264330$&$219$&$0.0008$\tabularnewline
6-10&6-17&European&$ 182937$&$ 80$&$0.0004$\tabularnewline
1-5&18-24&European&$ 107649$&$ 34$&$0.0003$\tabularnewline
6-10&18-24&European&$ 117840$&$ 36$&$0.0003$\tabularnewline
1-5&25+&European&$1001916$&$ 78$&$0.0001$\tabularnewline
6-10&25+&European&$ 724317$&$ 51$&$0.0001$\tabularnewline
1-5&0-2&Maori&$  10003$&$ 18$&$0.0018$\tabularnewline
6-10&0-2&Maori&$  30104$&$ 48$&$0.0016$\tabularnewline
1-5&3-5&Maori&$  10003$&$  7$&$0.0007$\tabularnewline
6-10&3-5&Maori&$  30104$&$ 11$&$0.0004$\tabularnewline
1-5&6-17&Maori&$  40461$&$ 19$&$0.0005$\tabularnewline
6-10&6-17&Maori&$ 116640$&$ 92$&$0.0008$\tabularnewline
1-5&18-24&Maori&$  15360$&$  5$&$0.0003$\tabularnewline
6-10&18-24&Maori&$  48495$&$  8$&$0.0002$\tabularnewline
1-5&25+&Maori&$  71217$&$  2$&$0.0000$\tabularnewline
6-10&25+&Maori&$ 192729$&$  6$&$0.0000$\tabularnewline
1-5&0-2&MLA&$    728$&$  0$&$0.0000$\tabularnewline
6-10&0-2&MLA&$   1290$&$  3$&$0.0023$\tabularnewline
1-5&3-5&MLA&$    728$&$  0$&$0.0000$\tabularnewline
6-10&3-5&MLA&$   1290$&$  1$&$0.0008$\tabularnewline
1-5&6-17&MLA&$   2991$&$  0$&$0.0000$\tabularnewline
6-10&6-17&MLA&$   4539$&$  1$&$0.0002$\tabularnewline
1-5&18-24&MLA&$   1710$&$  0$&$0.0000$\tabularnewline
6-10&18-24&MLA&$   3078$&$  2$&$0.0006$\tabularnewline
1-5&25+&MLA&$   8028$&$  0$&$0.0000$\tabularnewline
6-10&25+&MLA&$  10335$&$  6$&$0.0006$\tabularnewline
1-5&0-2&Pacific&$   2093$&$  5$&$0.0024$\tabularnewline
6-10&0-2&Pacific&$  13124$&$ 58$&$0.0044$\tabularnewline
1-5&3-5&Pacific&$   2093$&$  0$&$0.0000$\tabularnewline
6-10&3-5&Pacific&$  13124$&$  3$&$0.0002$\tabularnewline
1-5&6-17&Pacific&$   8541$&$  5$&$0.0006$\tabularnewline
6-10&6-17&Pacific&$  51183$&$ 22$&$0.0004$\tabularnewline
1-5&18-24&Pacific&$   3972$&$  1$&$0.0003$\tabularnewline
6-10&18-24&Pacific&$  22098$&$  8$&$0.0004$\tabularnewline
1-5&25+&Pacific&$  18492$&$  2$&$0.0001$\tabularnewline
6-10&25+&Pacific&$  91533$&$ 11$&$0.0001$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:percap}
\end{table}

For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used and present the results of a quasipoission regression model. We also account for differences in population sizes by using an offset term, the log(population size). We used a model simplification approach, by beginning our analyses with all terms and all interactions, and then simplifying the models through removal of non-significant higher order interaction terms. Thus, the final model that remained with all significant interaction terms had the following linear predictor:

\begin{equation} \label{eq:reg}
 log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a{}_e(x_a * x_e)+ log(population)  + \epsilon
  \end{equation}
Where $\alpha$ is the intercept, $y$ cases, $_a$ age, $_e$ Prioritized Ethnicity, $_n$ NZDep, and $\epsilon$ the error term.

@@

The distribution of the cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-014}
\end{center}
\caption{Distribution of cases per category used in the final regression model}
\label{fig:histcasecat}
\end{figure}

The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-015}
\end{center}
\caption{Regression model (Equation~\ref{eq:reg}) predictions plotted against the cases (Table~\ref{table:percap})}
\label{fig:predict}
\end{figure}

@@

A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}. The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented, though herd immunity of approximately 95 percent a homogeneously mixed immune population is recommended to interupt measles \citep{roberts0,roberts4}. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles. Thus, the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, estimated using the NZ census data \citep{stats14} and the proportion of each age class immune based on data provided to us by the Ministry of Health on commencing this project.

@@

\begin{equation} \label{eq:pc}
p_c = 1-(1/R_0)
  \end{equation}
To acocunt for the different outbreak sizes we weighted our $R_v$ estimates by outbreak size and used 1.96 standard deviations of the mean of our $R_v$ estimates, to generate 95\% confidence limits from 2009--2014 outbreaks or the 2013--2014 outbreak for our modelling scenarios below.

We then assume that the measles epidemics are occuring in a naive population that is the size of the naive and currently susceptible population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$. We estimated the susceptible population in NewZealand from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}). Thus, the $p_c$ applies to the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.

\begin{figure}[ht!]
\begin{center}
\includegraphics{interimreport2-021}
\end{center}
\caption{New Zealand population by age and estimated numbers of naive people in each age class}
\label{fig:naive}
\end{figure}

We then used the proportions of the population that were estimated to require vaccination ($p_c$, Equation~\ref{eq:pc}), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals. These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the benefit--cost analyses section below (Section~\ref{sub:cost_benefit}).

Finally, we used the same model structure and generation times to estimate the epidemic sizes once vaccination catch up protocols had been implemented assuming $R_v$ < 1 and the population of susceptible people in New Zealand was now reduced by the numbers vaccinated. Because the model is stochastic, we ran 1000 simulations and report the mean values, though the distribution of these is shown in the benefit--cost section (Section~\ref{sub:cost_benefit}).

@@

\begin{table}
\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
%latex.default(costtable3, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lrrr}
\hline\hline
\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Admitted}&\multicolumn{1}{c}{Proportion}\tabularnewline
\hline
\textless  15 months&$ 21$&$10$&$0.47$\tabularnewline
15 months – 3 years&$  7$&$ 1$&$0.14$\tabularnewline
4 – 9 years&$  8$&$ 0$&$0.00$\tabularnewline
10 – 19 years&$132$&$12$&$0.09$\tabularnewline
\textgreater  19 years&$ 33$&$11$&$0.33$\tabularnewline
Total&$201$&$34$&$0.17$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:freq}
\end{table}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-029}
\end{center}
\caption{Number of measles cases attending hospital and stay duration from 2000--2014}
\label{fig:hosp}
\end{figure}

@@

\begin{table}\small
\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2000--2014}
%latex.default(costtable4, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllrrl}
\hline\hline
\multicolumn{1}{c}{Year}&\multicolumn{1}{c}{Gender}&\multicolumn{1}{c}{Cost}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Length.of.stay}&\multicolumn{1}{c}{Cost.per.case}\tabularnewline
\hline
2000&F&4,296&$  2$&$  4$&2,148\tabularnewline
&M&4,554&$  4$&$  9$&1,139\tabularnewline
&Total&8,850&$  6$&$ 13$&1,475\tabularnewline
2001&F&3,740&$  5$&$  5$&748\tabularnewline
&M&7,527&$  8$&$ 13$&941\tabularnewline
&Total&11,267&$ 13$&$ 18$&867\tabularnewline
2002&F&924&$  2$&$  0$&462\tabularnewline
&M&2,945&$  3$&$  2$&982\tabularnewline
&Total&3,869&$  5$&$  2$&774\tabularnewline
2003&F&9,766&$  8$&$ 12$&1,221\tabularnewline
&M&475&$  1$&$  0$&475\tabularnewline
&Total&10,241&$  9$&$ 12$&1,138\tabularnewline
2004&F&1,437&$  1$&$  2$&1,437\tabularnewline
&M&3,328&$  3$&$  3$&1,109\tabularnewline
&Total&4,765&$  4$&$  5$&1,191\tabularnewline
2005&F&0&$  0$&$  0$&0\tabularnewline
&M&5,111&$  3$&$ 11$&1,704\tabularnewline
&Total&5,111&$  3$&$ 11$&1,704\tabularnewline
2006&F&0&$  0$&$  0$&0\tabularnewline
&M&602&$  1$&$  0$&602\tabularnewline
&Total&602&$  1$&$  0$&602\tabularnewline
2007&F&1,930&$  1$&$  3$&1,930\tabularnewline
&M&81,046&$  4$&$ 22$&20,262\tabularnewline
&Total&82,977&$  5$&$ 25$&16,595\tabularnewline
2008&F&714&$  1$&$  0$&714\tabularnewline
&M&2,324&$  2$&$  1$&1,162\tabularnewline
                    &Total&3,038&$  3$&$  1$&1,013\tabularnewline
2009&F&11,953&$  7$&$ 15$&1,708\tabularnewline
&M&28,830&$ 22$&$ 23$&1,310\tabularnewline
                    &Total&40,782&$ 29$&$ 38$&1,406\tabularnewline
2010&F&5,884&$  4$&$  5$&1,471\tabularnewline
&M&817&$  1$&$  0$&817\tabularnewline
                    &Total&6,701&$  5$&$  5$&1,340\tabularnewline
2011&F&103,460&$ 66$&$ 86$&1,568\tabularnewline
&M&101,842&$ 66$&$103$&1,543\tabularnewline
                    &Total&205,303&$132$&$189$&1,555\tabularnewline
2012&F&13,054&$  8$&$  6$&1,632\tabularnewline
&M&15,486&$ 11$&$  6$&1,408\tabularnewline
                    &Total&28,540&$ 19$&$ 12$&1,502\tabularnewline
2013&F&1,800&$  1$&$  2$&1,800\tabularnewline
&M&3,530&$  3$&$  4$&1,177\tabularnewline
                    &Total&5,330&$  4$&$  6$&1,333\tabularnewline
2014&F&55,633&$ 21$&$ 46$&2,649\tabularnewline
&M&77,014&$ 34$&$ 87$&2,265\tabularnewline
&Total&132,647&$ 55$&$133$&2,412\tabularnewline
2000-2014&      F&335,431&$166$&$284$&2,021\tabularnewline
&     M&214,591&$127$&$186$&1,690\tabularnewline
&   TOTAL&550,022&$293$&$470$&1,877\tabularnewline
\hline
\end{tabular}\end{center}\label{table:cases}
 \centering
 \begin{tablenotes}
      \small
      \item As of 7 August, 2014.
    \end{tablenotes}
\end{table}

@@

\subsection{Benefit--cost analyses methods}
To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{sec:epidemic_modelling}), we did several things. Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ distribution from (Section~\ref{sec:epidemic_modelling}) in the estimated susceptible population of naive New Zealanders (Figure~\ref{fig:naive}). We assumed recovery from infection led to immunity and thus constantly reduced the population size by the number recovered. We used these values of numbers of predicted cases and the cost figures above to estimate what the cost of not vaccinating additional populations would be. We also simulated what we might expect measles outbreaks to look like following introductions in the population, given that $R_v$ would be one, though where our estimates for $R_v$ are already < 1, we use this. We can therefore estimate the number of cases prevented, and the savings made from the additional vaccinations. We then use the costs of the catch up vaccination schemes, the costs of the expected measles-related costs due to constant introduction of measles despite increased population immunity, and the savings from reduced measles cases to work out the benefit--cost ratio ((B/C), where a B/C ratio > 1 means that the program benefits exceed their costs. A B/C value less than one suggests the costs are higher than the economic benefits. Lastly, benefits were assessed over a 10--year time period, using a discounting rate of 3\% discount per year for the costs, as is common for healthcare discounting \citep{honeycutt6}:
